---
subtitle: "TMA4268 Statistical Learning V2023"
title: "Compulsory Exercise 2: Title (give your project an informative title)"
author: "Torbjørn Vatne, Ludvik Braathen and Johan Bjerkem"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes: 
- \usepackage{amsmath}
- \usepackage{makecell}

output:
  # html_document:
  #   toc: no
  #   toc_depth: '2'
  #   df_print: paged
  pdf_document:
    toc: no
    toc_depth: '2'
urlcolor: blue
abstract: "**This is the place for your abstract (max 350 words).** Heart failure is one of the leading death causes in the world. Cardiovascular 
diseases (CVDs) which in many cases leads to heart failure stands for 
17.9 million deaths yearly. One third of people who suffer a heart attack or
a stroke die prematurely (under the age of 70). (https://www.who.int/health-topics/cardiovascular-diseases#tab=tab_1)

Detecting who is prone to heart failure is therefore an important task for 
society. First and foremost it's important as lives can be saved by early
detection and implementing counter measures. Secondly this is a huge cost to
our society and a burden on our hospitals."
---
  
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")

```

```{r,eval=TRUE,echo=FALSE}
#require("knitr")
require("rmarkdown")
require("ggthemes") #For themes on the plot, as the theme_excel
require("cowplot") #For the plotgrid function which makes several plots in one image
require("GGally")
library(knitr)
library(kableExtra)
library(dplyr)
library(scales)
require("tree")

require(tidyverse)
require(rlang)
```

<!--  Etc (load all packages needed). -->


## Introduction: Scope and purpose of your project

## Descriptive data analysis/statistics
### Load data
```{r, echo = TRUE, results='hide'}
heart <- read.csv("./heart.csv", header = T, stringsAsFactors = T)
str(heart)
summary(heart)
heart$FastingBS <- as.factor(heart$FastingBS)
heart$HeartDisease <- as.factor(heart$HeartDisease)
kable(head(heart), format = "markdown", align = "c", row.names = FALSE)
```
```{r, echo=FALSE}
truncate_colnames <- function(df, max_length = 10) {
  colnames(df) <- sapply(colnames(df), function(x) {
    if (nchar(x) > max_length) {
      paste0(substr(x, 1, max_length - 1), ".")
    } else {
      x
    }
  })
  return(df)
}

data <- heart
first_5_rows <- head(data, 5)

first_5_rows <- truncate_colnames(first_5_rows, max_length = 7)

kable(first_5_rows, format = "markdown", align = "c", row.names = FALSE)
```

We import the data set from file and immediately turn all columns with values
as strings into factors. After this we get a overview of the data 
using the *str* and *summary* methods. Here we notice that the values 
**fastingBS** and **heartDisease** need to be turned into factors. Lastly we
have a look at the head of the data.


### Histograms

```{r, fig.width=5}
tema = theme(plot.background = element_rect(fill="#e2e2e2"),
             plot.title = element_text(size=10, hjust=.5),
             axis.title.x = element_text(size=8, color = "black"),
             axis.text.x = element_text(size=6),
             axis.text.y = element_text(size=6))


hist_age <- ggplot(data = heart, mapping = aes(x = Age)) +
    geom_histogram(bins=25, fill = "red", color = "black", linewidth = 1.3) +
    theme_excel() +
    xlab("Age") +
    ylab("") +
    ggtitle("Age Histogram") +
    tema

hist_restbp <- ggplot(data = heart, mapping = aes(x = RestingBP)) +
    geom_histogram(bins=20, fill = "blue", color = "black", linewidth = 1.3) +
    theme_excel() +
    xlab("RestingBP") +
    ylab("") +
    ggtitle("Resting Blood Pressure Histogram") +
    tema

hist_chol <- ggplot(data = heart, mapping = aes(x = Cholesterol)) +
    geom_histogram(bins=20, fill = "green", color = "black", linewidth = 1.3) +
    theme_excel() +
    xlab("Cholesterol") +
    ylab("") +
    ggtitle("Cholesterol Histogram") +
    tema

hist_maxhr <- ggplot(data = heart, mapping = aes(x = MaxHR)) +
    geom_histogram(bins=20, fill = "pink", color = "black", linewidth = 1.3) +
    theme_excel() +
    xlab("Max HR") +
    ylab("") +
    ggtitle("Max Heart Rate Histogram") +
    tema

hist_maxhr_2 <- ggplot(data=heart, mapping=aes(MaxHR)) +
  geom_histogram(binwidth = 10, fill="pink") +
  xlab("Max HR") +
  ylab("Frequency") +
  ggtitle("Max Heart Rate") +
  theme_minimal()

plot_grid(hist_age, hist_restbp, hist_chol, hist_maxhr_2, nrow=2, ncol=2)
```

The **age** histogram seems legit and has no unexpected values. Both **resting 
blood pressure** and **cholesterol** seem to have some fishy 0 values. To our
knowledge and research these values are impossible, and seems to be a problem
with the data set. Continuing, the **max hr** plot also look okay, with no 
outliers.

#### Correlations continous variables

```{r}
heart$HeartDisease <- factor(heart$HeartDisease, levels = c(0, 1), labels = c("No", "Yes"))

ggplot(heart, aes(x = HeartDisease, y = Age, fill = HeartDisease)) +
  geom_violin() +
  labs(x = "Heart Disease",
       y = "Age") +
  theme_minimal()

ggplot(heart, aes(x = HeartDisease, y = RestingBP, fill = HeartDisease)) +
  geom_violin() +
  labs(x = "Heart Disease",
       y = "Resting Blood Pressure") +
  theme_minimal()

ggplot(heart, aes(x = HeartDisease, y = Cholesterol, fill = HeartDisease)) +
  geom_violin() +
  labs(x = "Heart Disease",
       y = "Cholesterol") +
  theme_minimal()

ggplot(heart, aes(x = HeartDisease, y = MaxHR, fill = HeartDisease)) +
  geom_violin() +
  labs(x = "Heart Disease",
       y = "MaxHR") +
  theme_minimal()

ggplot(heart, aes(x = Age, y = MaxHR)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "blue") +
  labs(x = "Age",
       y = "MaxHR") +
  theme_minimal()

ggplot(heart, aes(x = HeartDisease, y = Oldpeak, fill = HeartDisease)) +
  geom_violin() +
  labs(x = "Heart Disease",
       y = "Oldpeak") +
  theme_minimal()

ggcorr(heart)
#plot_grid(violin_age_dis, violin_restBP_dis, violin_chol_dis, violin_maxHR_dis, nrow=4, ncol=1)
```
People with heart diseases are significantly older on average. This is not too surprising, as older people are more prone to heart diseases as the human body gets worn out over time. There's somewhat correlation between Age and MaxHR, which makes sence since MaxHR generally decreases as a person gets older. This partially explains that people with heartdiseases generally have a lower MaxHR, however it might also be that people with heart diseases generally are less exposed to anaerobic training, which usually helps MaxHR stay higher.

#### Categorical variables

```{r}
categorical_plot <- function(df, cat_var1, cat_var2) {
  cat_var1_sym <- rlang::sym(cat_var1)
  cat_var2_sym <- rlang::sym(cat_var2)
  
  df_prop <- df %>%
    count(!!cat_var1_sym, !!cat_var2_sym) %>%
    group_by(!!cat_var2_sym) %>%
    mutate(Percentage = n / sum(n) * 100)
  
  plot <- ggplot(df_prop, aes(x = !!cat_var2_sym, y = Percentage, fill = !!cat_var1_sym)) +
    geom_bar(stat = "identity", position = "fill") +
    labs(x = cat_var2) +
    theme_minimal() +
    theme(legend.title = element_text(size = 4),
          legend.text = element_text(size = 4),
          aspect.ratio = 1) +
    scale_y_continuous(labels = percent_format())
  
  return(plot)
}

hd_sex <- categorical_plot(heart, "HeartDisease", "Sex")
hd_sex

hd_cpt <- categorical_plot(heart, "HeartDisease", "ChestPainType")
hd_cpt

cpt_sex <- categorical_plot(heart, "ChestPainType", "Sex")
cpt_sex

heart$FastingBS <- factor(heart$FastingBS, levels = c(0, 1), labels = c("No", "Yes"))

hd_fastingBS <- categorical_plot(heart, "HeartDisease", "FastingBS")
hd_fastingBS

hd_ECG <- categorical_plot(heart, "HeartDisease", "RestingECG")
hd_ECG

hd_exAng <- categorical_plot(heart, "HeartDisease", "ExerciseAngina")
hd_exAng

hd_ST <- categorical_plot(heart, "HeartDisease", "ST_Slope")
hd_ST

library(patchwork)
combined_plot <- (hd_sex | hd_cpt) /
                 (cpt_sex | hd_fastingBS) /
                 (hd_ECG | hd_exAng)

#print(combined_plot)
# This just squishes the plots together:
#plot_grid(hd_sex, hd_cpt, cpt_sex, hd_fastingBS, hd_ECG, hd_exAng, hd_ST, nrow=3, ncol=2)
```
In the first plot, we see that men are more likely to have a heart disease than women. We also see from the second plot that type of chest pain has a significant correlation with heart diseases. It is important to realize that many of the variables are correlated. For example, there are more asymptomatic men than women, the ChestPainType-category that is most likely for a person with a heart disease. As we see, ChestPainType depends on Sex. It is more difficult to see how resting electrocardiogram results predicts heart disease as all three categories are rather close in percentage of heart disease cases.

To summarize, many of the categorical variables seem to be correlated to heart disease and might therefore be valuable predictors, however one has to be aware that many of the variables are correlated within.

Error in cpt_sex | hd_fastingBS :
  operations are possible only for numeric, logical or comples types
Calls: <Anonymous> ... withVisible -> eval_with_user_handlers -> eval -> eval
Execution halted

## Methods
### Logistic Regression
```{r trainsplit}
set.seed(1)
train_perc <- 0.70
train_index <- sample(1:nrow(heart), nrow(heart)*train_perc)
test_index <- (-train_index)
train <- heart[train_index, ]
test <- heart[test_index, ]
```

```{r glm}
glm.fit <- glm(HeartDisease ~ ., data = train, family = binomial)
summary(glm.fit)

glm.probs <- predict(glm.fit, test,
type = "response")

glm.pred <- rep(0, length(glm.probs))
glm.pred[glm.probs > .5] <- 1

table(glm.pred, test$HeartDisease)
```
This model gives a good estimation when we set the cutoff range at 0.5.
```{r}
glm.fit2 <- glm(HeartDisease ~ . - Age - RestingBP - RestingECG - MaxHR, data = train, family = binomial)
summary(glm.fit2)

glm.probs2 <- predict(glm.fit2, test,
type = "response")

glm.pred2 <- rep(0, length(glm.probs2))
glm.pred2[glm.probs2 > .5] <- 1

table(glm.pred2, test$HeartDisease)
```
By looking at the summary of the previous model we found out at that we could remove **age**, **restingBP**, **restingEXg** and **MaxHR**.

## These can be tested: LDA, QDA, Naive bayes, KNN and Poisson. 

### Trees
```{r}
tree.fit <- tree(HeartDisease ~ ., data = train)
summary(tree.fit)

tree.pred <- predict(tree.fit, test, type = "class")
table(tree.pred, test$HeartDisease)
```
This performs almost as well as logistic regression.




## Results and interpretation


## Summary
